Manuscript_draft

### 2021-02-25 manuscript_draft.Rmd
manny work

A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p < 0.0001). Similar linear models fitted separately for fir and spruce showed that fir mortality had a similar relationship to plot basal area, whereas spruce mortality did not (not shown).  The diameter distribution of fir mortality followed a negative exponential distribution, with mortality concentrated in smaller diameter classes (`r figNums("mortSppDiam.plot",display="cite")`).  Spruce mortality was concentrated in larger diameter classes.  By treatment unit, the control unit had the highest mortality, whereas the high-removal treatment unit had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).  The majority of mortality occured during the most recent measurement period of 2009-2019.

@@

@@

In both species, post-treatment growth rates increased with the tree size at the time of treatment.  This has been found in other studies as well. and may be explained by the asymmetric competition theory whereby larger trees exert more competition on smaller trees. Implications of this result are that growth increases from treatment will be predominantly concentrated in the larger tree size classes.  If larger trees are able to increase their size faster than smaller trees and increase other metrics, such as crown size, to take advantage of increased growing space, then smaller trees will continue to experience competition and will not benefit as much from the thinning treatment.  Forest managers should account for potential size differential when planning thinning treatments.  For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.

@@

Stand density 
The high-removal treatment unit increased stand density more than the control unit over the 27-year study period and the differences between control and treatment units decreased in each subsequent measurement period.  In 2019, there were no significant stand density differences between all units.  These results suggest that the thinned stands increased density faster than the control and within three decades, there are no significant differences between the three units.  We note that these estimates are for trees with a minimum diameter of 7.5cm at breast height, and therefore our data do not account for new seedlings that may have become established over the time period of the study.  Thus, the phenomenon of increasing stand density may simply indicate that small trees responded with faster diameter growth in the thinned treatments (as previously discussed) and were able to achieve the minimum diameter class in this study. Increasing densities in the smaller diameter classes also suggests that these stands will maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.  Thus, thinning treatments may be used in this context to maintain a structure typical of uneven-aged stands.

The low-removal thinning treatment also maintained a species composition similar to the control, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.  In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.  Compared to spruce, subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.  Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.  This leads to a stand structure that is common in the Interior Region of BC and also found in the low-removal and control units.  The high-removal unit had a somewhat different species composition.  Here, fir was the leading species in all diameter classes.  This means that the high-removal thinning treatment converted the stand composition to fir-leading, which may be a primary concern for many forest managers, as spruce is often preferred as a crop species over subalpine fir [].  However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and spruce increased its composition compared to fir over the 27-year period of this study.  Over time, we expect that spruce would continue to increase its proportion relative to fir, resulting in similar species composition - diameter distributions as the other units.

Volume and basal area
Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that increment in these variables did not differ much between the two.  Thus, the long-term implications of this is that these stands are on similar development trajectories

@@

Stand mortality
Stand mortality increased with basal area after treatment.  This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.  Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.

>>>>

A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p < 0.0001). Mortality was generally not attributable to a particular forest health factor, and in the majority ofcases, we could not ascertain a casual agent.  Similar linear models fitted separately for fir and spruce showed that fir mortality had a similar relationship to plot basal area, whereas spruce mortality did not (not shown).  The diameter distribution of fir mortality followed a negative exponential distribution, with mortality concentrated in smaller diameter classes (`r figNums("mortSppDiam.plot",display="cite")`).  Spruce mortality was concentrated in larger diameter classes.  By treatment unit, the control unit had the highest mortality, whereas the high-removal treatment unit had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).  The majority of mortality occured during the most recent measurement period of 2009-2019.

@@

## Uneven-aged stand management
This study describes tree- and stand-level responses to partial harvest treatments in an uneven-aged stand.  
- THere are numerous studies documenting responses of even-aged stands to partial harvest treatments such as thinning.  In even-aged stands, a common objective of thinning treatments is to re-allocate growing space and resources to the most valulable trees in the stand that have capacity to increase production.  The objectives of partial harvesting in uneven-aged stands is more often to provide a sustained production of wood over multiple generations of trees.  In this context, partial harvesting in uneven-aged stands seeks to provide adequate growing space and resources to trees across a range of sizes and densities.  By quantifying growth responses at the tree- and stand-level, this study addresses questions regarding the manipulation of structure in an uneven-aged stand to achieve ...

@@

In both species, post-treatment growth rates increased with tree size at the time of treatment.  This has been found in other studies as well (references) and may be explained by the phenomenon of size-asymmetric competition whereby larger trees are able to preempt resources, especially light, from smaller individuals [@Bose2017].  The implications of size-assymetric competition in thinned stands is that size inequality among trees may increase [@Cordonnier2015], and growth increases from treatment will be predominantly concentrated in the larger tree size classes [@Bose2018; @Schwinning1998].  Smaller trees will likely continue to experience competition and will not benefit as much from the thinning treatment.  Thinning treatments may not promote growth responses in the largest docminant trees in a stand either, as these trees may not experience significant changes in resource availability or may have achieved maximal growth rates [@Ryan1997, Bose2017]. Forest managers should account for potential size differential when planning thinning treatments.  For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.

@@

Stand density 
The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the three units.  Increasing stand density reflects the growth of smaller trees that over time, achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.  The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).  We note that our study did not account for new seedlings that may have become established over the period of the study.  Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.  Thus, thinning treatments may be used in this context to maintain a structure typical of uneven-aged stands.

The low-removal thinning treatment also maintained a species composition similar to the unthinned stands, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.  This is a stand structure and composition common at low- to mid-elevations in the interior of British Columbia.  In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.  Compared to spruce, subalpine fir can be relatively short-lived, and can die after 250 years in the canopy.  Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.  The high-removal unit had a somewhat different species composition.  Here, fir was the leading species in all diameter classes.  This means that the high-removal thinning treatment converted the stand composition to fir-leading, which may be a primary concern for many forest managers, as spruce is often preferred as a crop species over subalpine fir [].  However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and over the 27-year period of this study, spruce monotonically increased its composition (by volume and basal area) compared to fir.  Over time, we expect that spruce would continue to increase its proportion relative to fir and eventually will achieve a similar species composition - diameter distribution as the other units.

Volume and basal area
Over the 27-year period of this study, the high-removal thinning treatments accrued basal area significantly faster than the control and low-removal thinning treatment, but not volume.  The volume formula used in this study uses both tree diameter and height [@Nigh2016].  Treatment-level differences in basal area increment but not volume suggest that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth. It can be difficult to precisely measure tree height growth using field-based methods, particularly in mature trees, which may be a primary reason that studies tend to focus on radial growth responses to silviculture treatments and gap dynamics [@Vepakommaetal2011].    Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].

Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that volume/basal area increment did not differ between the two.  The gaps created by the low-removal treatment may have been too small to elicit a strong growth response as observed in the high-removal thinning treatment.  Light levels may differ little between forest canopies that are closed or contain small gaps [@Bartemucci2002], and strong vertical light gradients may result in very little light regime changes for understorey trees.

Volume and basal area increment peaked in the 1994-1997 measurement period, with the strongest decline in annual growth over the last measurement period of 2009-2019.  Age-related declines in stand productivity may be associated with decreasing water-use efficiency and reduced photosynthesis, with the age of maximal growth rate varying with factors such as species and site conditions [@Ryan1997].  In both thinned treatments, maximal growth rates peaked in the 1997-2009 period.  Thus, it appears the thinning treatment may have delayed the period of maximal growth rates by approximately 10 years.  The timing of thinning treatments can be important.  Stands that are past their period of age-related maximal growth rate may not have as much capacity to respond to treatments as a younger stand, although this is liekly a more important factor for shade-intolerant species rather than the spruce and fir trees in this study [@Bose2017].

@@

Stand mortality
Stand mortality increased with basal area after treatment.  The thinned stands had the lowest mortality over the time period studies.  Thinning is considered a silviculture treatment that improves growth efficiency of the stand by redistibuting growing space and capturing mortality that would otherwise reduce net stand growth.  Our findings suggest that this occured, and that the thinning treament captured a portion of the stand that would have been lost eventually to natural mortality.  This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.  Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.

[@Valkonen2020] found similar mortality in uneven-aged norway spruce stands.
